---
title: "Spatial Transcriptomics Technology Comparison"
author: "Mariia Bilous"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")

# Same 3 environment variables every script in this package uses (see README.md). Knitting from
# RStudio's "Knit" button sets the working directory to this file's own directory, so the defaults
# below match code/lib.sh's own defaults (REPRO_ROOT = this directory) without needing to export
# anything first - but if you generated results with REPRO_DATA/REPRO_RESULTS pointing elsewhere,
# export them before knitting (or edit the three lines below) so this report finds them.
REPRO_ROOT <- Sys.getenv("REPRO_ROOT", unset = ".")
REPRO_DATA <- Sys.getenv("REPRO_DATA", unset = file.path(REPRO_ROOT, "data"))
REPRO_RESULTS <- Sys.getenv("REPRO_RESULTS", unset = file.path(REPRO_ROOT, "results"))
Sys.setenv(REPRO_ROOT = REPRO_ROOT, REPRO_DATA = REPRO_DATA, REPRO_RESULTS = REPRO_RESULTS)

source(file.path(REPRO_ROOT, "code/R/plot_common.R"))
suppressPackageStartupMessages(library(patchwork))

FIG_DIR <- file.path(REPRO_RESULTS, "figures")

# Every figure section below calls this once: show the real rendered PNG if 03_generate_figures.sh
# has already produced it (best fidelity - exact styling/dimensions); otherwise, if the lighter-
# weight CSV(s) behind it already exist (produced by 02_preprocess.sh, well before the plotting
# step), rebuild a faithful-but-unstyled version of the same plot on the fly; otherwise say so.
# This means this report can be knit at any point after 02_preprocess.sh, without waiting for
# 03_generate_figures.sh to run - it degrades gracefully to "what data exists right now."
show_or_rebuild <- function(png_path, csv_paths, rebuild_fn) {
  if (file.exists(png_path)) {
    knitr::include_graphics(png_path)
  } else if (length(csv_paths) > 0 && all(file.exists(csv_paths))) {
    cat("_(Rendered PNG not found - rebuilding a plain version from its underlying CSV(s).)_\n\n")
    print(rebuild_fn())
  } else if (length(csv_paths) == 0) {
    cat("_Not yet generated, and no CSV fallback is possible for this figure (see note below) -",
        "run the commands above first._\n")
  } else {
    missing <- csv_paths[!file.exists(csv_paths)]
    cat("_Not yet generated - missing:_\n\n", paste0("- `", missing, "`\n", collapse = ""), "\n",
        "_Run the commands above first._\n")
  }
}
```

## Overview

Reproduces the 8 figures from ["Reproducing the Atera vs Stratamap analysis"](https://lnkd.in/p/gHZ84RsX):
6 spatial transcriptomics technologies (**Atera**, **VisiumHD**, **Xenium MM/biomarkers**, **Xenium
v1/breast+100**, **Xenium 5k**, **StrataMap**) on the same breast cancer tissue block, all annotated
against the same 10x Chromium Flex scRNA-seq reference via RCTD (doublet mode, class-aware, Level2).

### Pipeline

1. **`00_download_raw_data.md`** - get the raw data.
2. **`02_preprocess.sh`** - raw data -> every intermediate artifact (QC, RCTD, embeddings, spatial
   metadata, spillover, composition/ARI/sensitivity summaries).
3. **`03_generate_figures.sh`** - the 8 PNGs below.
4. **This report** - narrates them, rendering from the PNG or its underlying CSV, whichever exists.

Each section names the `02_preprocess.sh` stage(s) and `03_generate_figures.sh` command for that
figure, then renders it: the real PNG if it exists, otherwise a version rebuilt from its CSV.

```{r shared-setup}
pal <- get_celltype_palette()
```

---

## Figure 1: Spatial distribution of cell types

One spatial scatter plot per sample (all 8, fixed order: StrataMap grades 1-3, VisiumHD, Atera,
Xenium 5k, Xenium MM, Xenium v1), points colored by RCTD `first_type`. This is the most direct look
at whether each technology recovers the tissue's known spatial organization (e.g. tumor regions,
stroma) at all, before any quantitative metric.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 (ingestion, QC, RCTD) then STAGE 5 (spatial metadata
extraction) for all 8 samples, then:
```bash
Rscript code/R/spatial_grid.R --out-dir results/figures/spatial --format png --dpi 300 \
  --point-size 0.3 --alpha 0.7 --panel-width 1.5 --panel-height 3 --out-name spatial_grid_all_bigdots_asp2
```

```{r fig1, fig.width=12, fig.height=3}
csvs <- vapply(TECH_ORDER, spatial_metadata_file_for_tech, character(1))
show_or_rebuild(
  file.path(FIG_DIR, "spatial/spatial_grid_all_bigdots_asp2.png"),
  csvs,
  function() {
    panels <- Filter(Negate(is.null), lapply(TECH_ORDER, function(t) build_spatial_plot(t, label_for_tech(t), pal)))
    wrap_plots(panels, nrow = 1)
  }
)
```

---

## Figure 2: Per-technology UMAP

Each technology's own UMAP (its full/native gene panel - "self", not restricted to any shared gene
list), colored by `first_type`. This shows whether cell-type structure is recoverable *in principle*
for each technology on its own terms, independent of any cross-technology gene restriction.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 then STAGE 4 (gene-set embeddings, "self"/`all` set)
for all 8 samples, then:
```bash
Rscript code/R/umap_by_geneset.R --out-dir results/figures/umap    # per-sample umap_self_<tech>.png
Rscript code/R/umap_self_grid.R --out-dir results/figures/umap --out-name umap_self_grid_all
```

```{r fig2, fig.width=12, fig.height=3}
show_or_rebuild(file.path(FIG_DIR, "umap/umap_self_grid_all.png"), character(0), NULL)
```

---

## Figure 3: Cell-type composition per sample

Cell-type proportions per sample, all 8 in one row, cell types ordered by their overall abundance
summed across every sample (same order in every panel, so panel-to-panel comparison is direct).

**To reproduce:** `02_preprocess.sh` STAGE 0-3, STAGE 5 (spatial metadata), and STAGE 7
(`celltype_composition.R` - writes `composition_summary.csv`, the CSV this figure and its fallback
both read), then:
```bash
Rscript code/R/composition_grid.R --summary-csv results/composition/composition_summary.csv \
  --out-dir results/figures/composition/sorted_by_abundance
```

```{r fig3, fig.width=12, fig.height=4}
composition_csv <- file.path(REPRO_RESULTS, "composition/composition_summary.csv")
show_or_rebuild(
  file.path(FIG_DIR, "composition/sorted_by_abundance/composition_grid_all_pct_linear_fixedy_tall.png"),
  composition_csv,
  function() {
    combined <- read.csv(composition_csv)
    overall_order <- combined %>% group_by(first_type) %>% summarise(total_n = sum(n)) %>%
      arrange(desc(total_n)) %>% pull(first_type)
    panels <- lapply(TECH_ORDER, function(t) {
      tab <- combined %>% filter(technology == t)
      if (nrow(tab) == 0) return(NULL)
      build_composition_plot(tab, label_for_tech(t), overall_order, pal, show_x_labels = FALSE, y_metric = "pct", y_scale = "linear")
    })
    wrap_plots(Filter(Negate(is.null), panels), nrow = 1)
  }
)
```

---

## Figure 4: Cell-type sensitivity to the nCount QC floor

For every sample, what % of each cell type's cells survive as the nCount QC threshold is raised well
above the pipeline's actual default (>=10) - up to 2000. Cell types with naturally low RNA content
(e.g. non-secretory, quiescent) lose a disproportionate share of their cells as the floor rises; this
figure makes that bias visible per technology.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 then STAGE 9 (`extract_count_distributions.R` -
writes the per-sample count-distribution parquets AND `celltype_threshold_sensitivity.csv`), then:
```bash
Rscript code/R/celltype_threshold_sensitivity.R --extract-dir results/count_distributions \
  --out-dir results/figures/threshold_sensitivity --layout row_tall
```

```{r fig4, fig.width=14, fig.height=2.5}
threshold_csv <- file.path(FIG_DIR, "threshold_sensitivity/celltype_threshold_sensitivity.csv")
show_or_rebuild(
  file.path(FIG_DIR, "threshold_sensitivity/celltype_threshold_sensitivity_tall.png"),
  threshold_csv,
  function() {
    combined <- read.csv(threshold_csv, stringsAsFactors = FALSE)
    combined$label <- factor(combined$label, levels = vapply(TECH_ORDER, label_for_tech, character(1)))
    combined$first_type <- factor(combined$first_type, levels = names(pal))
    ggplot(combined, aes(x = threshold, y = pct_retained, color = first_type)) +
      geom_line(linewidth = 0.6) + geom_point(size = 1) + scale_x_log10() +
      scale_color_manual(values = pal, drop = FALSE, name = "Cell type") +
      facet_wrap(~label, nrow = 1, scales = "free_y") +
      labs(x = "nCount threshold", y = "% cells retained") +
      theme_post_bar() + theme(legend.position = "none", strip.background = element_blank())
  }
)
```

---

## Figure 5: Self vs. shared-panel-genes summary

Five panels sharing one technology axis: N genes, N cells, N counts/cell, N counts/gene/cell, and
bioconservation (ARI between k-means-on-PCA clusters and RCTD `first_type`) - each comparing a
technology's own full panel ("self", solid) against the 353-gene panel shared across every
whole-transcriptome technology ("shared panel genes", hatched). This is the core "how much does
restricting everyone to the same gene axis change what you can resolve" figure.

Xenium 5k and Xenium MM don't have a *native* shared-panel-genes computation (their own panels
barely overlap the breast+100 gene list the shared axis is built from), so both reuse their own
`panel_genes` values as a stand-in - **except** Xenium 5k's hatched bar is intentionally omitted from
the nCount/nCount-per-nGenes/ARI panels (panels 3-5): restricting its cells to that narrow gene
subset filters out enough of them (`--min-count-in-subset`) that per-cell stats and clustering on the
remainder would be misleading. Xenium MM's panel (96 genes, all real signal, no filtering effect)
doesn't have this issue and keeps all 5 panels. N genes/N cells (panels 1-2) show both technologies'
proxy either way, since those two numbers stay honest regardless of filtering.

**To reproduce:** `02_preprocess.sh` STAGE 0-4 (embeddings) then STAGE 8 (separation metric/ARI),
STAGE 9 (count distributions), STAGE 10 (ngenes/ncells CSVs), and STAGE 11 (per-gene-set
sensitivity), then:
```bash
Rscript code/R/summary_grid_shared_panel.R \
  --ngenes-csv results/ngenes_ncells_barplot/ngenes_barplot.csv \
  --ncells-csv results/ngenes_ncells_barplot/ncells_barplot.csv \
  --overall-csv results/sensitivity/sensitivity_overall_all_technologies.csv \
  --ari-csv results/separation_ari/separation_ari_comparison.csv \
  --out-dir results/figures/summary_grid
```

```{r fig5, fig.width=11, fig.height=3.5}
summary_csv <- file.path(FIG_DIR, "summary_grid/summary_grid_data.csv")
show_or_rebuild(
  file.path(FIG_DIR, "summary_grid/summary_grid_shared_panel.png"),
  summary_csv,
  function() {
    d <- read.csv(summary_csv, stringsAsFactors = FALSE)
    d$label <- factor(d$label, levels = vapply(TECH_ORDER, label_for_tech, character(1)))
    d$gene_set <- factor(d$gene_set, levels = c("self", "shared_panel_genes"), labels = c("Self genes", "Shared panel genes"))
    panel_labs <- c(n_genes = "N genes", n_cells = "N cells", nCount = "N counts/cell",
                     nCount_per_nGenes = "N counts/gene/cell", ari = "Bioconservation (ARI)")
    d$panel <- factor(panel_labs[d$panel], levels = unname(panel_labs))
    ggplot(d, aes(x = label, y = value, fill = gene_set)) +
      geom_col(position = position_dodge(width = 0.75), width = 0.65) +
      facet_wrap(~panel, nrow = 1, scales = "free_x") +
      scale_fill_manual(values = c("Self genes" = "#4C72B0", "Shared panel genes" = "#55A868"), name = NULL) +
      coord_flip() +
      labs(x = NULL, y = NULL) +
      theme_post_bar() + theme(legend.position = "bottom")
  }
)
```

---

## Figure 6: RCTD secondary-type weight distribution

Per-sample distribution of RCTD's doublet-mode `weight_second_type` ("w2") - how much of each cell's
expression the model attributes to a second, distinct cell type. Higher/wider w2 distributions
suggest more segmentation-driven mixing of neighboring cells' transcripts, independent of any
spatial-neighborhood test (that's figure 7). Samples ordered by increasing median score.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 then STAGE 6 (`spillover.R` - writes the per-cell
spillover cache), then:
```bash
Rscript code/R/w2_boxplot.R --cache-root results/spillover_cache --out-dir results/figures/w2
```

```{r fig6, fig.width=6, fig.height=4}
w2_csv <- file.path(FIG_DIR, "w2/w2_boxplot_summary.csv")
show_or_rebuild(
  file.path(FIG_DIR, "w2/w2_boxplot_by_score.png"),
  w2_csv,
  function() {
    d <- read.csv(w2_csv, stringsAsFactors = FALSE)
    d$tech_family <- vapply(d$technology, tech_family, character(1))
    d$label <- factor(d$label, levels = rev(d$label[order(d$median)]))
    ggplot(d, aes(x = label, ymin = min, lower = p25, middle = median, upper = p75, ymax = max, fill = tech_family)) +
      geom_boxplot(stat = "identity", width = 0.6) +
      scale_fill_manual(values = TECH_FAMILY_COLORS, name = "Technology") +
      coord_flip() +
      labs(x = NULL, y = "secondary type weight (w2)") +
      theme_post_bar()
  }
)
```

---

## Figure 7: Spatial-neighborhood spillover (cosine similarity)

For every sample: cosine similarity between each cell's RCTD `weight_second_type` and its spatial
neighbors' pooled second-type weights, with a genuine permutation test (1000 shuffles of the
neighborhood term, one-sided p = fraction of shuffles >= observed, BH-corrected across the 8
samples). A high, significant cosine similarity means secondary-type signal is spatially structured
(neighbors "agree" on which second type is bleeding in) rather than random per-cell noise - evidence
that at least part of w2 (figure 6) reflects genuine spatial mixing/segmentation spillover, not just
per-cell annotation uncertainty.

**To reproduce:** same STAGE 6 spillover cache as figure 6, then:
```bash
Rscript code/R/spillover_metrics.R --cache-root results/spillover_cache --out-dir results/figures/spillover
```

```{r fig7, fig.width=5, fig.height=4}
spillover_csv <- file.path(FIG_DIR, "spillover/spillover_metrics_summary.csv")
show_or_rebuild(
  file.path(FIG_DIR, "spillover/spillover_cosine.png"),
  spillover_csv,
  function() {
    d <- read.csv(spillover_csv, stringsAsFactors = FALSE)
    d$tech_family <- vapply(d$technology, tech_family, character(1))
    d$label <- factor(d$label, levels = rev(d$label[order(d$cosine)]))
    ggplot(d, aes(x = label, y = cosine, fill = tech_family)) +
      geom_col() +
      scale_fill_manual(values = TECH_FAMILY_COLORS, name = "Technology") +
      coord_flip() +
      labs(x = NULL, y = "Cosine similarity") +
      theme_post_bar()
  }
)
```

---

## Figure 8: Per-gene sensitivity, own panel vs. shared gene axis

One point per technology: median transcripts/cell/gene on its own full panel (y) vs. on the 353-gene
shared panel axis (x), log-scaled, with a dashed identity line and IQR error bars in both directions.
Points above the identity line detect more signal per gene on their own panel than on the shared
axis; point size reflects panel size at self. This isolates whether a technology's sensitivity
advantage (or disadvantage) is about panel size/design specifically, independent of the cell-type
recovery metrics in the other figures.

**To reproduce:** `02_preprocess.sh` STAGE 0-3, STAGE 5 (spatial metadata, needed by
`gene_sensitivity.R`), and STAGE 12 (per-gene sensitivity + synthesis), then:
```bash
Rscript code/R/gene_sensitivity_scatter.R \
  --stats-csv results/gene_sensitivity/gene_sensitivity_stats.csv --out-dir results/figures/gene_sensitivity_scatter
```

```{r fig8, fig.width=6, fig.height=6}
scatter_csv <- file.path(FIG_DIR, "gene_sensitivity_scatter/scatter_self_vs_shared_panel_genes_transcripts_per_cell_median.csv")
show_or_rebuild(
  file.path(FIG_DIR, "gene_sensitivity_scatter/scatter_self_vs_shared_panel_genes_transcripts_per_cell_median_log10.png"),
  scatter_csv,
  function() {
    d <- read.csv(scatter_csv, stringsAsFactors = FALSE)
    d$short_label <- factor(d$short_label, levels = names(SAMPLE_COLORS))
    ggplot(d, aes(x = x, y = y)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
      geom_errorbar(aes(ymin = y_lo, ymax = y_hi, color = short_label), width = 0, alpha = 0.6) +
      geom_errorbarh(aes(xmin = x_lo, xmax = x_hi, color = short_label), height = 0, alpha = 0.6) +
      geom_point(aes(size = size, fill = short_label), shape = 21, color = "black") +
      scale_fill_manual(values = SAMPLE_COLORS, guide = "none") +
      scale_color_manual(values = SAMPLE_COLORS, guide = "none") +
      scale_size_continuous(name = "Genes in self panel", trans = "log10") +
      scale_x_log10() + scale_y_log10() +
      labs(x = "Shared panel genes (353), median transcripts/cell/gene",
           y = "Self-panel, median transcripts/cell/gene") +
      theme_bw() + theme(aspect.ratio = 1, legend.position = "bottom")
  }
)
```
