---
title: "Spatial Transcriptomics Technology Comparison"
author: "Mariia Bilous"
output:
  github_document:
    toc: true
    html_preview: false
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center",
                       fig.path = "report_figures/")

# Same 3 environment variables every script in this package uses (see README.md). Knitting from
# RStudio's "Knit" button sets the working directory to this file's own directory, so the defaults
# below match code/lib.sh's own defaults (REPRO_ROOT = this directory) without needing to export
# anything first - but if you generated results with REPRO_DATA/REPRO_RESULTS pointing elsewhere,
# export them before knitting (or edit the three lines below) so this report finds them.
REPRO_ROOT <- Sys.getenv("REPRO_ROOT", unset = ".")
REPRO_DATA <- Sys.getenv("REPRO_DATA", unset = file.path(REPRO_ROOT, "data"))
REPRO_RESULTS <- Sys.getenv("REPRO_RESULTS", unset = file.path(REPRO_ROOT, "results"))
Sys.setenv(REPRO_ROOT = REPRO_ROOT, REPRO_DATA = REPRO_DATA, REPRO_RESULTS = REPRO_RESULTS)

source(file.path(REPRO_ROOT, "code/R/plot_common.R"))
suppressPackageStartupMessages(library(patchwork))

FIG_DIR <- file.path(REPRO_RESULTS, "figures")

# Every figure section below calls this once: show the real rendered PNG if 03_generate_figures.sh
# has already produced it (best fidelity - exact styling/dimensions); otherwise, if that figure's own
# CSV exists (e.g. the PNG was deleted after a run, or figure 5/6/7/8's own script wrote its CSV but
# something failed before the PNG step), rebuild a faithful-but-unstyled version on the fly;
# otherwise say so. Figures 1 and 3 are the only ones whose fallback CSV comes straight from
# 02_preprocess.sh (spatial_metadata/*.csv, composition_summary.csv), so only those two degrade
# gracefully BEFORE 03_generate_figures.sh has ever run - figures 4-8's fallback CSVs are written by
# their own 03_generate_figures.sh scripts, so those specifically need that script to have run once.
show_or_rebuild <- function(png_path, csv_paths, rebuild_fn) {
  if (file.exists(png_path)) {
    if (isTRUE(knitr::pandoc_to("gfm"))) {
      # github_document needs a repo-relative image path it can commit alongside the .md -
      # REPRO_RESULTS commonly lives outside this repo (e.g. cluster scratch storage), so copy the
      # real rendered figure into report_figures/ instead of embedding that external path verbatim.
      local_dir <- file.path(REPRO_ROOT, "report_figures")
      dir.create(local_dir, showWarnings = FALSE)
      file.copy(png_path, file.path(local_dir, basename(png_path)), overwrite = TRUE)
      knitr::include_graphics(file.path("report_figures", basename(png_path)))
    } else {
      knitr::include_graphics(png_path)
    }
  } else if (length(csv_paths) > 0 && all(file.exists(csv_paths))) {
    cat("_(Rendered PNG not found - rebuilding a plain version from its underlying CSV(s).)_\n\n")
    print(rebuild_fn())
  } else if (length(csv_paths) == 0) {
    cat("_Not yet generated, and no CSV fallback is possible for this figure (see note below) -",
        "run the commands above first._\n")
  } else {
    missing <- csv_paths[!file.exists(csv_paths)]
    cat("_Not yet generated - missing:_\n\n", paste0("- `", missing, "`\n", collapse = ""), "\n",
        "_Run the commands above first._\n")
  }
}
```

## Overview

Reproduces the 8 figures from ["Reproducing the Atera vs Stratamap analysis"](https://lnkd.in/p/gHZ84RsX):
6 spatial transcriptomics technologies (**StrataMap**, **VisiumHD**, **Atera**, **Xenium MM/biomarkers**,
**Xenium v1/breast+100**, **Xenium 5k**) on breast cancer tissue - not one shared block, but largely
different samples entirely: Atera and Xenium MM/biomarkers are the only pair sharing a tissue block;
VisiumHD, Xenium v1/breast+100, and Xenium 5k are each their own separate specimen; and StrataMap
alone spans 3 different tumor grades (DCIS/IDC Grade 1-3), each its own distinct sample. Every cell,
regardless of which sample or disease stage it came from, is annotated against the same 10x Chromium
Flex scRNA-seq reference via [RCTD](https://doi.org/10.1038/s41587-021-00830-w) (Cable et al. 2022,
*Nature Biotechnology* - "Robust decomposition of cell type mixtures in spatial transcriptomics";
doublet mode, class-aware, Level2), run through [`rctd-py`](https://github.com/p-gueguen/rctd-py),
a public GPU-accelerated PyTorch reimplementation - not the original lab's private tooling.

### Pipeline

1. **`00_download_raw_data.md`** - get the raw data.
2. **`02_preprocess.sh`** - raw data -> every intermediate artifact (QC, RCTD, embeddings, spatial
   metadata, spillover, composition/ARI/sensitivity summaries).
3. **`03_generate_figures.sh`** - the 8 PNGs below.
4. **This report** - narrates them, rendering from the PNG or its underlying CSV, whichever exists.

Each section names the `02_preprocess.sh` stage(s) and `03_generate_figures.sh` command for that
figure, then renders it: the real PNG if it exists, otherwise a version rebuilt from its CSV.

```{r shared-setup}
pal <- get_celltype_palette()

# Figures 1-4 all color by RCTD first_type but hide their own per-panel legend by design (grid
# assembly, no room for 8 repeated legends) - this draws the shared color key once per figure below.
draw_celltype_legend <- function() {
  legend_plot <- ggplot(data.frame(cell_type = factor(names(pal), levels = names(pal))),
                         aes(x = 1, y = 1, color = cell_type)) +
    geom_point(size = 5) +
    scale_color_manual(values = pal, name = "Cell type", drop = FALSE) +
    guides(color = guide_legend(nrow = 2, override.aes = list(size = 5))) +
    theme_void() +
    theme(legend.position = "bottom", legend.title = element_text(size = 12, face = "bold"),
          legend.text = element_text(size = 11))
  g <- ggplotGrob(legend_plot)
  grob_names <- vapply(g$grobs, function(x) x$name, character(1))
  legend_grob <- g$grobs[[which(grepl("guide-box", grob_names))[1]]]
  grid::grid.newpage()
  grid::grid.draw(legend_grob)
}
```

---

## Figure 1: Spatial distribution of cell types

One spatial plot per sample (all 8, fixed order: StrataMap grades 1-3, VisiumHD, Atera,
Xenium 5k, Xenium MM, Xenium v1), points colored by RCTD `first_type`. This is the most direct look
at whether each technology recovers the tissue's spatial organization (e.g. tumor regions,
stroma) at all, before any quantitative metric - and, since these are largely different samples (see
Overview), it also demonstrates just how different and heterogeneous the tissue architecture itself
is from one specimen to the next, independent of any technology effect.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 (ingestion, QC, RCTD) then STAGE 5 (spatial metadata
extraction) for all 8 samples, then:
```bash
Rscript code/R/spatial_grid.R --out-dir results/figures/spatial --format png --dpi 300 \
  --point-size 0.3 --alpha 0.7 --panel-width 1.5 --panel-height 3 --out-name spatial_grid_all_bigdots_asp2
```

```{r fig1, fig.width=12, fig.height=3}
csvs <- vapply(TECH_ORDER, spatial_metadata_file_for_tech, character(1))
show_or_rebuild(
  file.path(FIG_DIR, "spatial/spatial_grid_all_bigdots_asp2.png"),
  csvs,
  function() {
    panels <- Filter(Negate(is.null), lapply(TECH_ORDER, function(t) build_spatial_plot(t, label_for_tech(t), pal)))
    wrap_plots(panels, nrow = 1)
  }
)
```

```{r fig1-legend, fig.width=10, fig.height=1.2}
draw_celltype_legend()
```

---

## Figure 2: Per-technology UMAP

Each technology's own UMAP (its full/native gene panel - "self", not restricted to any shared gene
list), colored by `first_type`. This shows whether cell-type structure is recoverable *in principle*
for each technology on its own terms, independent of any cross-technology gene restriction.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 then STAGE 4 (gene-set embeddings, "self"/`all` set)
for all 8 samples, then:
```bash
Rscript code/R/umap_by_geneset.R --out-dir results/figures/umap    # per-sample umap_self_<tech>.png
Rscript code/R/umap_self_grid.R --out-dir results/figures/umap --out-name umap_self_grid_all
```

```{r fig2, fig.width=12, fig.height=3}
show_or_rebuild(file.path(FIG_DIR, "umap/umap_self_grid_all.png"), character(0), NULL)
```

```{r fig2-legend, fig.width=10, fig.height=1.2}
draw_celltype_legend()
```

---

## Figure 3: Cell-type composition per sample

Cell-type proportions per sample, cell types ordered by their overall abundance
summed across every sample (same order in every panel).

**To reproduce:** `02_preprocess.sh` STAGE 0-3, STAGE 5 (spatial metadata), and STAGE 7
(`celltype_composition.R` - writes `composition_summary.csv`, the CSV this figure and its fallback
both read), then:
```bash
Rscript code/R/composition_grid.R --summary-csv results/composition/composition_summary.csv \
  --out-dir results/figures/composition/sorted_by_abundance
```

```{r fig3, fig.width=12, fig.height=4}
composition_csv <- file.path(REPRO_RESULTS, "composition/composition_summary.csv")
show_or_rebuild(
  file.path(FIG_DIR, "composition/sorted_by_abundance/composition_grid_all_pct_linear_fixedy_tall.png"),
  composition_csv,
  function() {
    combined <- read.csv(composition_csv)
    overall_order <- combined %>% group_by(first_type) %>% summarise(total_n = sum(n)) %>%
      arrange(desc(total_n)) %>% pull(first_type)
    panels <- lapply(TECH_ORDER, function(t) {
      tab <- combined %>% filter(technology == t)
      if (nrow(tab) == 0) return(NULL)
      build_composition_plot(tab, label_for_tech(t), overall_order, pal, show_x_labels = FALSE, y_metric = "pct", y_scale = "linear")
    })
    wrap_plots(Filter(Negate(is.null), panels), nrow = 1)
  }
)
```

```{r fig3-legend, fig.width=10, fig.height=1.2}
draw_celltype_legend()
```

---

## Figure 4: Cell-type sensitivity to the nCount QC floor

For every sample, what % of each cell type's cells survive as the nCount QC threshold is raised well
above the pipeline's actual default (>=10) - up to 2000. Cell types with naturally low RNA content lose a disproportionate share 
of their cells as the floor rises; this figure makes that bias visible per technology.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 then STAGE 9 (`extract_count_distributions.R` -
writes the per-sample count-distribution parquets this figure reads; `celltype_threshold_sensitivity.csv`
itself is written by the plotting command below, not by STAGE 9), then:
```bash
Rscript code/R/celltype_threshold_sensitivity.R --extract-dir results/count_distributions \
  --out-dir results/figures/threshold_sensitivity --layout row_tall
```

```{r fig4, fig.width=14, fig.height=2.5}
threshold_csv <- file.path(FIG_DIR, "threshold_sensitivity/celltype_threshold_sensitivity.csv")
show_or_rebuild(
  file.path(FIG_DIR, "threshold_sensitivity/celltype_threshold_sensitivity_tall.png"),
  threshold_csv,
  function() {
    combined <- read.csv(threshold_csv, stringsAsFactors = FALSE)
    combined$label <- factor(combined$label, levels = vapply(TECH_ORDER, label_for_tech, character(1)))
    combined$first_type <- factor(combined$first_type, levels = names(pal))
    ggplot(combined, aes(x = threshold, y = pct_retained, color = first_type)) +
      geom_line(linewidth = 0.6) + geom_point(size = 1) + scale_x_log10() +
      scale_color_manual(values = pal, drop = FALSE, name = "Cell type") +
      facet_wrap(~label, nrow = 1, scales = "free_y") +
      labs(x = "nCount threshold", y = "% cells retained") +
      theme_post_bar() + theme(legend.position = "none", strip.background = element_blank())
  }
)
```

```{r fig4-legend, fig.width=10, fig.height=1.2}
draw_celltype_legend()
```

---

## Figure 5: Self vs. shared-panel-genes summary

Five panels sharing one technology axis: N genes, N cells, N counts per cell, N counts/gene per cell, and
bioconservation (ARI between k-means-on-PCA clusters and RCTD `first_type`) - each comparing a
technology's own full panel ("self", solid) against the 353-gene panel shared across every
whole-transcriptome technology ("shared panel genes", hatched). This is the core "how much does
restricting everyone to the same gene axis change what you can resolve" figure.

Xenium 5k and Xenium MM don't have a *native* shared-panel-genes computation, so their hatched bar
is a proxy: their own genes intersected with the 380-gene breast+100 addon panel (`panel_genes`).
For Xenium MM this shrinks its 280-gene panel down to 96 genes; for Xenium 5k,
its ~5,100-gene panel down to 241. **Except** Xenium 5k's hatched bar is intentionally
omitted from the nCount/nCount-per-nGenes/ARI panels (panels 3-5): restricting its cells to that
241-gene subset filters out enough of them (`--min-count-in-subset`) that per-cell stats and
clustering on the remainder would be misleading. Xenium MM's 96-gene subset (all real signal, no
filtering effect) doesn't have this issue and keeps all 5 panels. N genes/N cells (panels 1-2) show
both technologies' proxy either way, since those two numbers stay honest regardless of filtering.

**To reproduce:** `02_preprocess.sh` STAGE 0-4 (embeddings) then STAGE 8 (separation metric/ARI),
STAGE 9 (count distributions), STAGE 10 (ngenes/ncells CSVs), and STAGE 11 (per-gene-set
sensitivity), then:
```bash
Rscript code/R/summary_grid_shared_panel.R \
  --ngenes-csv results/ngenes_ncells_barplot/ngenes_barplot.csv \
  --ncells-csv results/ngenes_ncells_barplot/ncells_barplot.csv \
  --overall-csv results/sensitivity/sensitivity_overall_all_technologies.csv \
  --ari-csv results/separation_ari/separation_ari_comparison.csv \
  --out-dir results/figures/summary_grid
```

```{r fig5, fig.width=11, fig.height=3.5}
summary_csv <- file.path(FIG_DIR, "summary_grid/summary_grid_data.csv")
show_or_rebuild(
  file.path(FIG_DIR, "summary_grid/summary_grid_shared_panel.png"),
  summary_csv,
  function() {
    d <- read.csv(summary_csv, stringsAsFactors = FALSE)
    d$label <- factor(d$label, levels = vapply(TECH_ORDER, label_for_tech, character(1)))
    d$gene_set <- factor(d$gene_set, levels = c("self", "shared_panel_genes"), labels = c("Self genes", "Shared panel genes"))
    panel_labs <- c(n_genes = "N genes", n_cells = "N cells", nCount = "N counts/cell",
                     nCount_per_nGenes = "N counts/gene/cell", ari = "Bioconservation (ARI)")
    d$panel <- factor(panel_labs[d$panel], levels = unname(panel_labs))
    ggplot(d, aes(x = label, y = value, fill = gene_set)) +
      geom_col(position = position_dodge(width = 0.75), width = 0.65) +
      facet_wrap(~panel, nrow = 1, scales = "free_x") +
      scale_fill_manual(values = c("Self genes" = "#4C72B0", "Shared panel genes" = "#55A868"), name = NULL) +
      coord_flip() +
      labs(x = NULL, y = NULL) +
      theme_post_bar() + theme(legend.position = "bottom")
  }
)
```

---

## Figure 6: RCTD secondary-type weight distribution

Per-sample distribution of RCTD's doublet-mode `weight_second_type` ("w2"), which represents the
fraction of each cell's expression profile attributed to a second, distinct cell type. Higher w2
values indicate greater profile impurity, as a larger proportion of the observed expression is
explained by a different cell type. Samples are ordered by increasing median w2.

**To reproduce:** `02_preprocess.sh` STAGE 0-3 then STAGE 6 (`spillover.R` - writes the per-cell
spillover cache), then:
```bash
Rscript code/R/w2_boxplot.R --cache-root results/spillover_cache --out-dir results/figures/w2
```

```{r fig6, fig.width=6, fig.height=4}
w2_csv <- file.path(FIG_DIR, "w2/w2_boxplot_summary.csv")
show_or_rebuild(
  file.path(FIG_DIR, "w2/w2_boxplot_by_score.png"),
  w2_csv,
  function() {
    d <- read.csv(w2_csv, stringsAsFactors = FALSE)
    d$tech_family <- vapply(d$technology, tech_family, character(1))
    d$label <- factor(d$label, levels = rev(d$label[order(d$median)]))
    ggplot(d, aes(x = label, ymin = min, lower = p25, middle = median, upper = p75, ymax = max, fill = tech_family)) +
      geom_boxplot(stat = "identity", width = 0.6) +
      scale_fill_manual(values = TECH_FAMILY_COLORS, name = "Technology") +
      coord_flip() +
      labs(x = NULL, y = "secondary type weight (w2)") +
      theme_post_bar()
  }
)
```

---

## Figure 7: Spatial-neighborhood spillover (cosine similarity)

For every sample: cosine similarity between each cell's RCTD `weight_second_type` and its spatial
neighbors' pooled second-type weights, with a genuine permutation test (1000 shuffles of the
neighborhood term, one-sided p = fraction of shuffles >= observed, BH-corrected across the 8
samples). A high, significant cosine similarity means secondary-type signal is spatially structured
(neighbors "agree" on which second type is bleeding in) rather than random per-cell noise - evidence
that at least part of w2 (figure 6) reflects genuine spatial mixing/segmentation spillover, not just
per-cell annotation uncertainty.

**To reproduce:** same STAGE 6 spillover cache as figure 6, then:
```bash
Rscript code/R/spillover_metrics.R --cache-root results/spillover_cache --out-dir results/figures/spillover
```

```{r fig7, fig.width=5, fig.height=4}
spillover_csv <- file.path(FIG_DIR, "spillover/spillover_metrics_summary.csv")
show_or_rebuild(
  file.path(FIG_DIR, "spillover/spillover_cosine.png"),
  spillover_csv,
  function() {
    d <- read.csv(spillover_csv, stringsAsFactors = FALSE)
    d$tech_family <- vapply(d$technology, tech_family, character(1))
    d$label <- factor(d$label, levels = rev(d$label[order(d$cosine)]))
    ggplot(d, aes(x = label, y = cosine, fill = tech_family)) +
      geom_col() +
      scale_fill_manual(values = TECH_FAMILY_COLORS, name = "Technology") +
      coord_flip() +
      labs(x = NULL, y = "Cosine similarity") +
      theme_post_bar()
  }
)
```

---

## Figure 8: Per-gene sensitivity, own panel vs. shared gene axis

One point per technology: median transcript abundance (i.e., for each gene, total gene counts
normalized by total number of cells) on the technology's own full panel (y) vs. on the shared-panel
axis (x), log-scaled, with a dashed identity line and IQR error bars in both directions. Point size
reflects self panel size.

The shared-panel axis is the same literal 353-gene list for every technology, but not every
technology's own panel actually contains all 353 of those genes: Atera, VisiumHD, Xenium v1, and
StrataMap (all 3 grades) do (353/353); Xenium 5k's panel only overlaps 229 of them, and Xenium MM's
smaller targeted panel only 91. Those two points are still the correct, literal intersection with
the same 353-gene list - just a smaller one, because their panels are smaller.

**To reproduce:** `02_preprocess.sh` STAGE 0-3, STAGE 5 (spatial metadata, needed by
`gene_sensitivity.R`), and STAGE 12 (per-gene sensitivity + synthesis), then:
```bash
Rscript code/R/gene_sensitivity_scatter.R \
  --stats-csv results/gene_sensitivity/gene_sensitivity_stats.csv --out-dir results/figures/gene_sensitivity_scatter
```

```{r fig8, fig.width=6, fig.height=6}
scatter_csv <- file.path(FIG_DIR, "gene_sensitivity_scatter/scatter_self_vs_shared_panel_genes_transcripts_per_cell_median.csv")
show_or_rebuild(
  file.path(FIG_DIR, "gene_sensitivity_scatter/scatter_self_vs_shared_panel_genes_transcripts_per_cell_median_log10.png"),
  scatter_csv,
  function() {
    d <- read.csv(scatter_csv, stringsAsFactors = FALSE)
    d$short_label <- factor(d$short_label, levels = names(SAMPLE_COLORS))
    ggplot(d, aes(x = x, y = y)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
      geom_errorbar(aes(ymin = y_lo, ymax = y_hi, color = short_label), width = 0, alpha = 0.6) +
      geom_errorbarh(aes(xmin = x_lo, xmax = x_hi, color = short_label), height = 0, alpha = 0.6) +
      geom_point(aes(size = size, fill = short_label), shape = 21, color = "black") +
      scale_fill_manual(values = SAMPLE_COLORS, guide = "none") +
      scale_color_manual(values = SAMPLE_COLORS, guide = "none") +
      scale_size_continuous(name = "Genes in self panel", trans = "log10") +
      scale_x_log10() + scale_y_log10() +
      labs(x = "Shared panel genes (353), median transcripts/cell/gene",
           y = "Self-panel, median transcripts/cell/gene") +
      theme_bw() + theme(aspect.ratio = 1, legend.position = "bottom")
  }
)
```
